---
title: "Group Assessment 1"
author: "430189999"
date: "data here"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: "yeti"
---

###Load Libraries
```{r load_libraries,results="hide",message=FALSE}
##loading data

library(readxl)
library(dplyr)
library(tibble)
library(ggplot2)
library(stringr)
library(readr)
library(crayon)
library(questionr)
library(ggpubr)

#for fancy ggpairs plot
#library(GGally)

#for clustering on catgeorical variables
#library(FactoMineR)

#easy regression table?
#library(stargazer)

#nice looking tables
library(kableExtra)

```
###Load Data
```{r}
df <- read_csv("https://raw.githubusercontent.com/mantissa-aidan/amed3002/master/All_sample_data.csv",
    col_types = cols(CEBPA_Biallelic = col_factor(levels = c("y", 
        "n")), CumulativeChemo = col_factor(levels = c("y", 
        "n"))))
```

###Domain Information

##Data Dictionary
```{r}
dataDictionary <- read_delim("https://raw.githubusercontent.com/mantissa-aidan/amed3002/master/dataDictionary.csv", 
    "\t", escape_double = FALSE, col_names = FALSE, 
    trim_ws = TRUE)

dataDictionary %>%
  kable() %>%
  kable_styling()
```

##Helper Functions
```{r}
###REARRANGE COLS
##arrange df vars by position
##'vars' must be a named vector, e.g. c("var.name"=1)
arrange.vars <- function(data, vars){
    ##stop if not a data.frame (but should work for matrices as well)
    stopifnot(is.data.frame(data))

    ##sort out inputs
    data.nms <- names(data)
    var.nr <- length(data.nms)
    var.nms <- names(vars)
    var.pos <- vars
    ##sanity checks
    stopifnot( !any(duplicated(var.nms)), 
               !any(duplicated(var.pos)) )
    stopifnot( is.character(var.nms), 
               is.numeric(var.pos) )
    stopifnot( all(var.nms %in% data.nms) )
    stopifnot( all(var.pos > 0), 
               all(var.pos <= var.nr) )

    ##prepare output
    out.vec <- character(var.nr)
    out.vec[var.pos] <- var.nms
    out.vec[-var.pos] <- data.nms[ !(data.nms %in% var.nms) ]
    stopifnot( length(out.vec)==var.nr )

    ##re-arrange vars by position
    data <- data[ , out.vec]
    return(data)
}

#USAGE
#table <- data.frame(Time=c(1,2), In=c(2,3), Out=c(3,4), Files=c(4,5))
#table
##  Time In Out Files
##1    1  2   3     4
##2    2  3   4     5

#arrange.vars(table, c("Out"=2))
##  Time Out In Files
##1    1   3  2     4
##2    2   4  3     5

#arrange.vars(table, c("Out"=2, "Files"=1, "Time"=4))
##  Files Out In Time
##1     4   3  2    1
##2     5   4  3    2


```

###Cleaning
```{r, results = 'asis'}
df$VitalStatus
df$VitalStatus[df$VitalStatus == "Healthy, pooled CD34+"] <- "Alive"
df$VitalStatus[df$VitalStatus == "Healthy, Individual BM MNC"] <- "Alive"
df$VitalStatus[df$VitalStatus == "Healthy, Individual CD34+"] <- "Alive"
df$VitalStatus[df$VitalStatus == "0.9"] <- "Unknown"

ggplot(df, aes(VitalStatus, as.numeric(df$WBC_Count), fill = Gender)) + geom_boxplot()

df$CumulativeTreatmentTypeCount <- as.numeric(df$CumulativeTreatmentTypeCount)
df$OverallSurvival <- as.numeric(df$OverallSurvival)
df$PlateletCount <- as.numeric(df$PlateletCount)
df$TimeOfCollectionRelativeToInclusion <- as.numeric(df$TimeOfCollectionRelativeToInclusion)
df$WBC_Count <- as.numeric(df$WBC_Count)

c <- unique(df$PriorMalignancyType)
c

colnames(df)
df <- arrange.vars(df, c("CumulativeTreatmentTypeCount"=1, "OverallSurvival"=2, "PlateletCount"=3))
df <- arrange.vars(df, c("TimeOfCollectionRelativeToInclusion"=4, "WBC_Count"=5))

glimpse(df)
factors = c(6:48)
df[factors] <- lapply(df[factors], factor)
df <- na.omit(df)

dataContinuous = as.data.frame(model.matrix(~.,df)[,-1])
pca = prcomp(dataContinuous)
summary(pca)$importance[,1:6]
plot(pca$x[,1:2])
sort(pca$rotation[,1])
```

##Tab Heading {.tabset .tabset-fade .tabset-pills}

#Our Content

###Fatima
```{r}
#Code goes here
survival rate and Percenatge of Blast in bone marrow

Adult acute myeloid leukemia is cancer which affects the bone marrow and produces abnormal myeloblasts. it is a cancer of the blood and bone marrow which creates excess white blood cells, red blood cells, or platelets. 


does the percentage of blast in the bone marrow increase the white blood cell count?
```{r}
ggplot(df, aes(x = factor(PercentBlastsInBM), y = df$WBC_Count)) + geom_boxplot()+labs(x = "% OF BLAST IN BONE MARROW", y = "WBC COUNT")
```
the graph above demonstrates the amount of white blood cell present in patients compared to their percentage of blast in their bone marrow. it can be seen that the higher the perecentage of blast is, the more white blood cells were counted.  


what is the survival rate of patients who recorded a higher % of blast in the bone marrow?
is there a correlation between the % of blast present in bone marrow and patients overall survival rate?

```{r}
summary(myData$data.OverallSurvival)
OverallSurvivalNum <-as.numeric(myData$data.OverallSurvival)
mean(OverallSurvivalNum) 
str(OverallSurvivalNum)
summary(OverallSurvivalNum)
```

the average survival rate collect in this data is 323 days, the higesr being 5305 day and the lowest at -1 day.

```{r}
plot(data$PercentBlastsInBM,myData$data.OverallSurvival)
```
The graph above demonstrates the percentage of blast present in the bone marrow of patients along with their overall survival. It can be seen that no correlation between both factors is present as are seenn to have very similar survival regardless to the perecent of blast present in their bone marrow.



is there a correlation present between white blood cell count, percentage of blast in the bone marrow and the overall survival of patients?

```{r}
interaction.plot(x.factor = df$OverallSurvival, trace.factor = df$PercentBlastsInBM, 
                 response = df$WBC_Count, fun = mean, 
                 type = "b", legend = TRUE, 
                 xlab = "Overall survival", ylab="WBC count",
                 pch=c(1,10), col = c("#00AFBB", "#E7B800"))
```

conclusion
With the graphs conducted, it can be concluded that the perecentage of blast in the bone marrow does not place an affect on the overall survival rate, however it does place an affect on the white blood cell count
```

###Yousra

####Does relapse of disease indicate lower chances of survival?

####Hypothesis: Relapse decreases odds of survival

```{r}
#Code goes here
#What is Leaukeuamia, specifically ALM?
```
####Acute Myeloid Leukemia results specifically from muations in developing bone marrow cells. This leads to unrestrained proliferation of leukemic white blood cells usually accompanied by anemia, impaired blood clotting, and enlargement of the lymph nodes, liver, and spleen.

####What factors (prior diagnosis, age at diagnosis, gender) contribute to Overall survival
```{r}
#How many rows and columns are in the data?
nrow(df)
ncol(df)
#How many of the subjects were Female?
sum(df$Gender == "Female")
#How many were Male?
sum(df$Gender == "Male")
```

####Relationship between relapse and vitality status

```{r}
#How many subjects had a prior diagnosis of MPN?
PriorMDSpositive <- sum(df$PriorMDS == "y")
summary(PriorMDSpositive)
#How many subjects who had a prior diagnosis of MPN are dead?
dead <- sum(df$VitalStatus == "Dead")
summary(dead)
sum(dead=PriorMDSpositive)
```
#### 100% of subjects who had been priorly diagnosed with the disease have been confirmed dead. 


####Relationship between relapse and overall survival
```{r}
#### Prior MDS/Malignancy on Overall Survival {.tabset .tabset-fade}

##### PLot

``{r}

bold.italic.blue = element_text(face = "bold.italic", colour = "blue")

ggplot(df, aes(df$PriorMalignancyNonMyeloid, df$OverallSurvival,fill=df$PriorMDS)) + 
  geom_boxplot() + 
  labs(title = "Duration of survival from AML after prior malgnancy", x = "Prior Maglignany", y="Survival (Days after Diagnosis)", fill="Prior \nMyelodysplastic \nsyndrom \ndiagnosis") +
  theme( axis.title.x = bold.italic.blue, axis.title.y = bold.italic.blue, plot.title = element_text( hjust=0.5, vjust=1, face='bold.italic', colour = "blue"), panel.border = element_rect(linetype = "dashed", fill = NA))


```


##### Analysis

```{r}

SurvivaLAOV=aov(df$OverallSurvival ~ df$PriorMalignancyNonMyeloid * df$PriorMDS)

summary(SurvivaLAOV)

Taby=table(df$PriorMalignancyNonMyeloid,  df$VitalStatus)

chisq.test(Taby)


```

<font size="1"> **Figure .** **Effects of prior morbidities on Survival** This figure demonstrates whether survival (in days) after diagnosis of AML is affected by previous individual history of cancer (p > 0.05) and whether their is a relationship between prior malignancy and vital status in follow up (p < 0.05) and the OR in this relationship (p < 0.0005). Assumptions that the data is normally distributed and is independant have been satisfied.</font>

Although the data on figure 2 demonstrates that the difference between surival (in days) based on whether you had a prior malignancy or not is non-significant (One-Way Anova Analysis), a chi-squared analysis deteremined that there was a relationship between prior malignancy and whether the patient would be alive when it was time for follow up (p < 0.05). This was further solidified with an Odds Ratio analysis, which determined with 95% confidence that 97.5% of people with prior malignancies will not survive until the follow up date.




```
####Figure 1.  demonstrates whether survival (in days) after diagnosis of AML is affected by previous diagnonsis





###Vinitha
#Understanding the risk factors of leukemia

### What is leukemia?
#### leukemia is a type of cancer which develops in blood - forming tissues such as the bone marrow 
#### It is usually caused by high numbers of white blood cells in the circulation 
#### Hence testing for white blood cell counts could be useful in the diagnosis of leukemia
#### Although, it is not directly used to diagnose cancer it can be used to indicate the need for further testing 

### Statistics of leukemia
#### Over 60, 000 Australians are currently living with leukemia 
#### At least 13, 000 more Australians are newly diagnosed with leukemia each year 
#### Around 4,900 of these individuals pass away due to leukemia each year 

### AIM
#### To raise awareness about the risk factors of leukemia, to allow people in risk groups to undergo frequent testing for leukaemia. This will allow individuals to be diagnosed earlier, possibly decreasing death rates

### Target audience
#### It is targeted to the general public 
#### Leukemia is a disease that affects white blood cells in individuals resulting in abnormally high white blood cell counts. Hence testing for white blood cell counts could be useful in the diagnosis of leukemia. Although, it is not directly used to diagnose cancer it provides information on whether further testing is needed.


#### Question: Does gender affect white blood cell count in individuals which potentially indicate a risk of leukemia?

#### Hypothesis: Studies have found that males have a higher risk of leukemia which means they could have a higher white blood cell count compared to females 
#### H0 = There is no difference in white blood cell count between males and females 
#### H1 = There is a difference in white blood cell count between males and females 


#### Supporting studies: A study performed northeast peninsular Malaysia, found that acute leukemia is more common in males possibily due to differences in "sex-responsive" genes near the ABO gene locus at chromosome 9. 

#### Reference: Jackson, N., Menon, B.S., Zarina, W., Zawawi, N. and Naing, N.N., 1999. Why is acute leukemia more common in males? A possible sex-determined risk linked to the ABO blood group genes. Annals of hematology, 78(5), pp.233-236.

#### Analysis of data - comparing gender and white blood cell count 
```{r}

ggplot(df, aes(x = factor(Gender), y = df$WBC_Count, fill = Gender)) + geom_boxplot()+labs(x = "GENDER", y = "WBC COUNT (x10^9 cells/litre)")

```
#### Figure 1: Boxplot comparing white blood cell counts between males and females.


#### From figure 1, it can be observed that both boxplots are right skewed which means mean > median value. A greater median value as well as a greater interquartile range is seen in females compared to males. This indicates that females have a higher white blood cell count compared to males. However, males are observed to have more outliers present. 

#### Violin plot is also used to show the probability density of data at different values

```{r}

ggplot(df, aes(x = factor(Gender), y = df$WBC_Count, fill = Gender)) + geom_violin()+labs(x = "GENDER", y = "WBC COUNT (x10^9 cells/litre)")
```
#### Figure 2: Violin plot comparing white blood cell counts between males and females.

#### Figure 2 shows that males have a higher distribution of white blood cell count compared to females. It is also seen that females have a higher probabilty of having a white blood cell count of over 50 (x10^9 cells/litre).



#### Conclusion: Results from the figures indicate that females have a higher white blood cell count compared to males. Hence, there is weak evidence against the null hypothesis. However, it must be noted that having a higher white blood cell count does not directly suggest a greater risk of developing leukemia in individuals. It is still important to monitor white blood cell counts to check for abnormally high levels which indicate the need for further testing. 


#### Checking for further risk factors - Myelodysplastic syndromes(MDS)

#### MDS are a group of diseases which affect the production of normal blood cells in the bone marrow 

 
#### Question: Does MDS affect white blood cell count differently in males and females?
#### Hypothesis: 
#### H0 = MDS does not affect white blood cell count differently in males and females 
#### H1 - MDS affects white blood cell count differently in males and females

#### Two-way ANNOVA test 
#### Assumption: Observations within each cell are normally distributed and have equal variances

```{r}
res.aov2 <- aov(df$WBC_Count ~ df$Gender + df$PriorMDS, data = df)
summary(res.aov2)
```
#### Level of significance = 0.05 

#### Pr (>F) value is > 0.05, there is weak evidence against the null hypothesis (H0)

####Not the above fitted model is called additive model. It makes an assumption that the two factor variables are independent.

#### Check if these two variables might interact to create an synergistic effect

#### Two-way ANOVA with interaction effect
#### These two calls are equivalent

```{r}
res.aov3 <- aov(df$WBC_Count ~ df$Gender * df$PriorMDS, data = df)
res.aov3 <- aov(df$WBC_Count ~ df$Gender + df$PriorMDS + df$Gender:df$PriorMDS, data = df )
            
summary(res.aov3)
```
#### Pr(>F) value is > 0.05, there is weak evidence against the null hypothesis. 


#### Interaction plot for two way ANNOVA test
```{r}
interaction.plot(x.factor = df$PriorMDS, trace.factor = df$Gender, 
                 response = df$WBC_Count, fun = mean, 
                 type = "b", legend = TRUE, 
                 xlab = "MDS", ylab="WBC count",
                 pch=c(1,19), col = c("#00AFBB", "#E7B800"))
```


#### Conclusion: From the ANOVA table we can conclude that both prior MDS and gender or their interaction are not statistically significant. These results would lead us to believe that being previously diagnosed with MDS or difference in gender, will not significantly impact the white blood cell count of individuals.


### David 

#### Relationship between MDS and WBC Count {.tabset .tabset-fade}

##### MDS Vs WBC count

```{r}

MDSplot1=ggplot(df,aes(x=df$PriorMDS,y=df$WBC_Count, fill=df$PriorMDS)) + geom_boxplot() + labs(title = "MDS Diagnosis effect on WBC count", x = "Diagnosed with \nMyelodysplastic Syndrome", y="WBC Count \n(Cells/L of blood)") + theme(legend.position = "none")

MDSplot2=ggplot(df,aes(x=df$PriorMDS,y=df$WBC_Count, fill=df$PriorMalignancyNonMyeloid)) + geom_boxplot() + labs(title = "MDS Diagnosis effect on WBC count", x = "Diagnosed with \nMyelodysplastic Syndrome", y="WBC Count \n(Cells/L of blood)", fill="Prior \nNon Myeloid \nNeoplasm")

MDSplot1





```


#####MDS + PriorNOn... vs WBC

```{r}

MDSplot2

```


##### Statistical Analysis


```{r}

aov(df$WBC_Count ~ df$PriorMalignancyNonMyeloid * df$PriorMDS)

summary(aov(df$WBC_Count ~ df$PriorMalignancyNonMyeloid * df$PriorMDS))

```

####  WBC on Overall Survival {.tabset .tabset-fade}

##### WBC v overall Survival

```{r}
df4 = df[df$OverallSurvival < 2000,]

LMFIT = lm(df$OverallSurvival ~ df$WBC_Count)
LMFIT2 = lm(df4$OverallSurvival ~ df4$WBC_Count)

LMPLOT1=ggplot(df,aes(x =df$WBC_Count, y=df$OverallSurvival )) +geom_point() + theme_bw() + geom_smooth(method = 'lm') + labs(x = "WBCs/L", y="Survial (Days)")

LMPLOT2=ggplot(df4,aes(x =df4$WBC_Count, y=df4$OverallSurvival )) +geom_point() + theme_bw() + geom_smooth(method = 'lm') + labs(x = "WBCs/L", y="Survial (Days)")

ggarrange(LMPLOT1, LMPLOT2, labels = c("A", "B")) 
```

##### WBC v Survival Linear regression analysis

```{r}
summary(LMFIT)
summary(LMFIT2)

```

##### WBC v Vitality status

```{r}
WBCplot3=ggplot(df,aes(x=df$VitalStatus,y=df$WBC_Count, fill=df$VitalStatus)) + geom_boxplot() + labs(title = "WBC v Vitality status on check up", x = "Survival Until Check up date", y="WBC Count \n(Cells/L of blood)") + theme(legend.position = "none")

WBCplot3

```

##### WBC v Vitality analysis

```{r}

WBCvVITALITYaov=aov(df$WBC_Count ~ df$VitalStatus)

summary(WBCvVITALITYaov)

```



###Aidan
```{r}
#Code goes here
```

###Tab1

```{r exploration_plotting}
#ggplot(df, aes(x=df$`Age at Dx (years)`, fill = df$`Age at Dx (years)`)) +
#  geom_bar(position = "dodge2") +
#    labs(title = "Age/Gender distribution",  fill = "Age", x = "Gender", y = "Count")
```

###Tab 2

```{r}
#both distrbutions are approx normal
#ggplot(df, aes(x=df$Gender, fill = df$`Age at Dx (years)`)) +
#  geom_bar(position = "dodge2") +
#    labs(title = "Age/Gender distribtion",  fill = "Age", x = "Gender", y = "Count")
```

###Tab 3


```{r}
#ggplot(df, aes(x=df$Race, fill = df$`Age at Dx (years)`)) +
#  geom_bar(position = "dodge2") +
#    labs(title = "Race/Gender included in study", x = "Gender", y = "Count", fill = "Age")
```

##Tab Section 2{.tabset .tabset-fade .tabset-pills}

###Tab 2.1

```{r}
#ggplot(df, aes(x = df$Gender, y = df$`Survival (months)`, fill = df$Gender)) + geom_boxplot() +
#facet_wrap(~ df$Disease, ncol = 50) +
#  labs(title = "Survival time between men and women, by disease", x = "Gender", y = "Survival Time")
```

###Tab 2.2

```{r}
##ggplot(df,aes(x = df$`Age at Dx (years)`, y = df$`Survival (months)`, colour = df$Disease)) +
#  geom_point() +
#  geom_jitter(height = 0.05) +
#  labs(colour = "Disease", y="Months Survived", x="Age", title="Comparing Age and Survival by Disease type") +
#  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10))
```


###Tab 2.3

```{r}
#ggplot(df,aes(x=df$`Age at Dx (years)`, y=df$`Survival (months)`, colour = df$`OnStudy Therapy Radiation Type`, #shape = df$`OnStudy Therapy Chemo Agent Name`)) +
#  geom_point() +
#  facet_wrap(~df$Disease) +
#  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7)) +
#  labs(shape = "patient had chemo", colour = "patient had RT", y="Age at diagnosis",x="Months survived" ,title = "Comparing age and survival time by treatment type, between diseases")
```

##Analysis

```{r analysis}

```

```{r}
```

```{r}

```




