---
title: "Assessment2_Works"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

## We want to reproduce these results

Twenty genes displayed AML-specific expression changes that were not found in the normal hematopoietic cells. Subsequent analyses using microarray data from 285 additional AML patients confirmed expression changes for 13 of the 20 genes. Seven genes (BIK, CCNA1, FUT4, IL3RA, HOMER3, JAG1, WT1) displayed increased expression in AML, while 6 genes (ALDHA1A, PELO, PLXNC1, PRUNE, SERPINB9, TRIB2) displayed decreased expression. 

Quantitative RT/PCR studies for the 7 over-expressed genes were performed in an independent set of 9 normal and 21 pediatric AML samples. All 7 over-expressed genes displayed an increased expression in the AML samples compared to normals. Three of the 7 over-expressed genes (WT1, CCNA1, and IL3RA) have already been linked to leukemogenesis and/or AML prognosis, while little is known about the role of the other 4 over-expressed genes in AML.

```{r}
#BiocManager::install("GEOquery")
#biocLite("GEOquery")

library(Biobase)
library(GEOquery)
library(dplyr)

library(edgeR)
library(limma)
#these are some genes that were particularly highly expressed
#we'll try to find them using DEA
high_result <- c("BIK", "CCNA1", "FUT4", "IL3RA", "HOMER3", "JAG1", "WT1")
#low expressed results from study
low_result <- c("ALDHA1A", "PELO", "PLXNC1", "PRUNE", "SERPINB9", "TRIB2")
#Load in the soft file
gds3057 <- getGEO(filename='GDS3057_full.soft.gz')

#naming some stuff to use later
disease <- gds3057@dataTable@columns

head(disease)

df <- gds3057@dataTable@table
eset <- df

#Subsetting just the important expression data
eset <- eset[,3:66]

#There is a key/value relationship between the genes
#Donm't know why, it results it duplicate genes being recorded
rname <- df$ID_REF

#renamingrows of expression set
rownames(eset) <- rname

#dropping unncessary info
#renaming rows of disease state
dRowNames <- disease$sample
disease <- disease %>% select(-sample)
disease <- disease %>% select(-cell.type,-description)
rownames(disease) <- dRowNames

#disease looks like this now
head(disease)

#construct design matrix
#One hot encoding
design <- model.matrix(~disease.state, disease)

#looks like this
head(design)
head(eset)

#We can fit a linear model
fit <- lmFit(eset, design)
#Do heaps of T-tests
fit <- eBayes(fit, robust = TRUE)

summary(fit)

#we can see which genes were high and lowly expressed by this model
results <- decideTests(fit[, "disease.statenormal"])
summary(results)


#Now we can do the same thing with DGE, see if we get better results
dge <- DGEList(counts = eset[, rownames(design)])
dge <- calcNormFactors(dge)

#Some form of variance stabilisation
v <- voom(dge, design, plot = TRUE)

#Fitting the linear model with the voom data
fit <- lmFit(v, design)
fit <- eBayes(fit, robust = TRUE)

#See how results differ from previous testing method
results <- decideTests(fit[, "disease.statenormal"])
summary(results)

#I'm a bit worried about this histogram
#I think it should be uniform, but there's loads of 0's
hist(fit$p.value[, 2])

#where our p values sit around 0
volcanoplot(fit, coef = "disease.statenormal")

#A list of the top genes
topGenes = topTable(fit, coef = "disease.statenormal", n = 1e+06, p.value = 0.05)

########
#PROPER RESULTS HERE

#here are the pvalues
topGenes
check_final <- rownames(topGenes)

#These are the features with the lowest p values
top_10_lowest_pvalues_ids <- check_final[1:10]

top_10_lowest_pvalues_ids

pvals <- topGenes$P.Value

#we have to look up which features correspond to which specific genes
Id_to_gene <- df %>% filter(df$ID_REF %in% top_10_lowest_pvalues_ids)

#you can see which id corresponds to which gene here
Id_to_gene

high_result

low_result


results2 <- df[,1:2]

results2

results3 <- results2 %>% filter(df$ID_REF %in% top_10_lowest_pvalues)

results3

check2 <- df


eset




```

```{r}

library(Biobase)
library(GEOquery)
library(dplyr)

gds3057 <- getGEO(filename='GDS3057_full.soft.gz')


glimpse(gds3057)

colnames(Table(gds3057))

Meta(gds3057)$feature_count
Meta(gds3057)$sample_count

Table(gds3057)[1:100,1:6]


tablecols <- gds3057@dataTable@columns

head(tablecols)

disease_state <- tablecols$disease.state

df <- gds3057@dataTable@table

dim(df)

head(df)

#table(duplicated(df$IDENTIFIER))
#table(duplicated(df$ID_REF))

df2 <- df %>% filter(IDENTIFIER == "CCNA1")

df2 <- df2[,3:66]
 
df2
ccna_exp <- df2[1,]
ccna_exp

what <- as.data.frame(cbind(t(ccna_exp)))

what

ccna <- (t(df2))
colnames(ccna) <- c("CCNA Expression Level")

ccna <- cbind(ccna, disease_state)

ccna[,1]
plot(ccna[,1])

ccna <- as.data.frame(ccna)
ccna

library(ggplot2)

ggplot(ccna, aes(rownames(ccna),ccna$`CCNA Expression Level`, colour = disease_state)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  ggtitle("Expression level of CNNA amongst patients with/without leakumia")

ccna

head(ccna)

#ident <- df[,2]
id_ref <- df[,1]

colnames(df3) <- id_ref

head(df3)


head(eset)
countdata <- as.matrix(eset)

head(countdata)

# Get log2 counts per million
logcounts <- log2(countdata + 1)

# make a colour vector
statusCol <- as.numeric(factor(disease_state)) + 1
# Check distributions of samples using boxplots
b <- boxplot(logcounts, 
        xlab="", 
        ylab="Log2(Counts)",
        las=2,
        col=statusCol)


library(ggfortify)
#BiocManager::install("DESeq2")
library(DESeq2)
rlogcounts <- rlog(countdata)
# run PCA
pcDat <- prcomp(t(countdata))
# plot PCA
autoplot(pcDat)


sort(pcDat$rotation[1:10,1])


# setting shape to FALSE causes the plot to default to using the labels
autoplot(pcDat,
         data = disease, 
         colour="disease.state", 
         shape=FALSE,
         label.size=2)


# We estimate the variance for each row in the logcounts matrix
countVar <- apply(rlogcounts, 1, var)
# Get the row numbers for the top 500 most variable genes
highVar <- order(countVar, decreasing=TRUE)[1:500]
# Subset logcounts matrix
hmDat <- rlogcounts[highVar,]

```




