---
title: "LabReport3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("https://bioconductor.org/biocLite.R")

# Load the BiocInstaller package
library(BiocInstaller)

# Check R version
version

# Explicit syntax to check the Bioconductor version
BiocInstaller::biocVersion()

library(BSgenome)

packageVersion("BSgenome")


```

#looking at yeast as an example
```{r}
#BiocManager::install("BSgenome.Scerevisiae.UCSC.sacCer3")
yeastGnome <- library(BSgenome.Scerevisiae.UCSC.sacCer3)

```

limma course from DataCamp
```{r}


```



## We want to reproduce these results

Twenty genes displayed AML-specific expression changes that were not found in the normal hematopoietic cells. Subsequent analyses using microarray data from 285 additional AML patients confirmed expression changes for 13 of the 20 genes. Seven genes (BIK, CCNA1, FUT4, IL3RA, HOMER3, JAG1, WT1) displayed increased expression in AML, while 6 genes (ALDHA1A, PELO, PLXNC1, PRUNE, SERPINB9, TRIB2) displayed decreased expression. 

Quantitative RT/PCR studies for the 7 over-expressed genes were performed in an independent set of 9 normal and 21 pediatric AML samples. All 7 over-expressed genes displayed an increased expression in the AML samples compared to normals. Three of the 7 over-expressed genes (WT1, CCNA1, and IL3RA) have already been linked to leukemogenesis and/or AML prognosis, while little is known about the role of the other 4 over-expressed genes in AML.


```{r}

## We'll load a count, samplesheet and clinical data file

#load(url("http://www.maths.usyd.edu.au/u/ellisp/AMED3002/melanoma.RData"))

dim(counts)

#the data is very large

```

```{r}
library(ggplot2)
library(ggfortify)
library(rlang)
library(dplyr)

counts2 = counts[rowMeans(counts) > 20, ]
counts2 <- counts2[, !duplicated(colnames(counts2))]
```

If we plot the first two principle components we see they acccount for 15% and 5.2% of variation.

We can see theres does appear to be some structure.
```{r}
# Perform PCA
pca = prcomp(t(counts2), scale = TRUE)

autoplot(pca)

sort(pca$rotation[1:10,1])
sort(pca$rotation[1:10,2])
```
Further scaling and plotting the first 3 three pc's reveals some apparent structure between each pc.

We can try to see a relationship between the pca's an primary diagnoses but there are many levels of diagnosis.
So if there is some relationship it's unclear.

```{r}
cpm1 = apply(counts2, 2, function(x) x/sum(x)) * 1e+06

pca1 = prcomp(t(cpm1), scale = TRUE)

pairs(pca1$x[, 1:3], col = as.numeric(clinical$primary_diagnosis))
pal = colorRampPalette(c("white", "black"))(ncol(counts))
pairs(pca1$x[, 1:3], col = pal[rank(colSums(counts))])

sort(pca1$rotation[1:10,1])
sort(pca1$rotation[1:10,2])

```

Once we adjust for multiple multiple hypothesis testing we can see the relationship becomes stronger.
```{r}
library(edgeR)
dge = DGEList(counts2)
x2 <- calcNormFactors(dge, method = "TMM")
cpm2 <- cpm(x2)
cpm2 = cpm2[rownames(counts2), ]

pca2 = prcomp(t(cpm2), scale = TRUE )

pairs(pca2$x[, 1:3], col = as.factor(clinical$primary_diagnosis))
pal = colorRampPalette(c("blue", "red"))(ncol(counts2))
pairs(pca2$x[, 1:3], col = pal[rank(colSums(counts2))])


```




```{r}
library(edgeR)
library(limma)

```

I'm not a biologist so I don't understand how to relate these genes I've found to some larger library of genes. I'm not even sure what pathways are. Pathways are something like a series of genes that produce a phenotype?

How dO I find the path ways of the genes I found in line 106?
What codes/naming convetions should I be matching up, searching for?


```{r}
library(biomaRt)

```

```{r}


```
Trying to perform a fishers test but I can't get the pathways

```{r}

```






