---
title: "Assessment2_Works"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#BiocManager::install("GEOquery")
library(tibble)
library(tidyverse)
source("https://bioconductor.org/biocLite.R")
biocLite("GEOquery")
library(Biobase)
library(GEOquery)
library(dplyr)
library(edgeR)
library(limma)
library(ggfortify)
#BiocManager::install("DESeq2")
library(DESeq2)
```

## R Markdown

## We want to reproduce these results

Twenty genes displayed AML-specific expression changes that were not found in the normal hematopoietic cells. Subsequent analyses using microarray data from 285 additional AML patients confirmed expression changes for 13 of the 20 genes. Seven genes (BIK, CCNA1, FUT4, IL3RA, HOMER3, JAG1, WT1) displayed increased expression in AML, while 6 genes (ALDHA1A, PELO, PLXNC1, PRUNE, SERPINB9, TRIB2) displayed decreased expression. 

Quantitative RT/PCR studies for the 7 over-expressed genes were performed in an independent set of 9 normal and 21 pediatric AML samples. All 7 over-expressed genes displayed an increased expression in the AML samples compared to normals. Three of the 7 over-expressed genes (WT1, CCNA1, and IL3RA) have already been linked to leukemogenesis and/or AML prognosis, while little is known about the role of the other 4 over-expressed genes in AML.

```{r}



#We can fit a linear model
fit <- lmFit(eset, design)
#Do heaps of T-tests
fit <- eBayes(fit, robust = TRUE)

summary(fit)

#we can see which genes were high and lowly expressed by this model
results <- decideTests(fit[, "disease.statenormal"])
summary(results)


#Now we can do the same thing with DGE, see if we get better results
dge <- DGEList(counts = eset[, rownames(design)])
dge <- calcNormFactors(dge)

#Some form of variance stabilisation
v <- voom(dge, design, plot = TRUE)

#Fitting the linear model with the voom data
fit <- lmFit(v, design)
fit <- eBayes(fit, robust = TRUE)

#See how results differ from previous testing method
results <- decideTests(fit[, "disease.statenormal"])
#summary(results)

#I'm a bit worried about this histogram
#I think it should be uniform, but there's loads of 0's
binw <- 0.1
hist(fit$p.value[, 2], by = binw)

#where our p values sit around 0
volcanoplot(fit, coef = "disease.statenormal", highlight = 12, names = df$IDENTIFIER)

#A list of the top genes
topGenes = topTable(fit, coef = "disease.statenormal", n = 1e+06, p.value = 0.05)

########
#PROPER RESULTS HERE

#here are the pvalues
topGenes
check_final <- rownames(topGenes)

#These are the features with the lowest p values
top_10_lowest_pvalues_ids <- check_final[1:10]

top_10_lowest_pvalues_ids

pvals <- topGenes$P.Value

#we have to look up which features correspond to which specific genes
Id_to_gene <- df %>% filter(df$ID_REF %in% top_10_lowest_pvalues_ids)

Id_to_gene

#you can see which id corresponds to which gene here
#Id_to_gene

#up_regulated

#down_regulated

ggplot(fit$pvalue, aes(x = fit$p.value[,2])) +
  geom_histogram(binwidth = 0.001)


```

#Matching up our genes to our ID's
```{r}

#Pulling out only the genes we found to be significant
joinMe = NULL
joinMe$ID_REF <- rownames(topGenes)
joinMe$pval <- topGenes$adj.P.Val

joinMe <- as.data.frame(joinMe)

joinMe2 <- as.data.frame(df[,1:2])

#head(joinMe)
#head(joinMe2)

results <- inner_join(joinMe,joinMe2, by = "ID_REF")
#results



```
## This is all plotting

```{r}

#Checking phenotype of a gene we found to be significant

df2 <- df %>% filter(IDENTIFIER == "ALDH1A1")

df2 <- df2[,3:66]
 
#df2
ccna_exp <- df2[1,]
#ccna_exp

#what <- as.data.frame(cbind(t(ccna_exp)))

#what

ccna <- (t(df2))
colnames(ccna) <- c("CCNA Expression Level")

ccna <- cbind(ccna, disease$disease.state)

#ccna[,1]
plot(ccna[,1])

ccna <- as.data.frame(ccna)

library(ggplot2)
#library(tidyverse)

ggplot(ccna, aes(rownames(ccna),ccna$`CCNA Expression Level`, colour = disease$disease.state)) +
  geom_point(xlab= "Samples", ylab="Expression Level") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  ggtitle("Expression level of ALDH1A1 amongst patients with/without leakumia")

#ident <- df[,2]
id_ref <- df[,1]



# make a colour vector

```
```{r}
statusCol <- as.numeric(factor(disease$disease.state)) + 1
# Check distributions of samples using boxplots
boxplot(countdata, 
        xlab="", 
        ylab="Expression Value",
        las=2,
        col=statusCol,
        main = "Distribution of expression level per sample")
```


```{r}
#countdata
# run PCA
pcDat <- prcomp(t(countdata))
# plot PCA
autoplot(pcDat)


sort(pcDat$rotation[1:10,1])


# setting shape to FALSE causes the plot to default to using the labels
autoplot(pcDat,
         data = disease, 
         colour="disease.state", 
         shape=FALSE,
         label.size=2,
         main = "Dimension Reduction: First two Principal components")

# We estimate the variance for each row in the logcounts matrix
countVar <- apply(countdata, 1, var)
# Get the row numbers for the top 500 most variable genes
highVar <- order(countVar, decreasing=TRUE)[1:1000]
# Subset logcounts matrix
hmDat <- countdata[highVar,]

library(gplots)
library(RColorBrewer)
# Get some nicer colours
mypalette <- brewer.pal(11, "RdYlBu")
# http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3
morecols <- colorRampPalette(mypalette)

#levels(disease$disease.state)

# Set up colour vector for celltype variable
col.cell <- c("purple","orange")[disease$disease.state]
# Plot the heatmap
heatmap.2(hmDat, 
          col=rev(morecols(50)),
          trace="column", 
          main="Top 1000 most variable genes across samples",
          ColSideColors=col.cell,scale="row")

```








